A tool set for random number generation on GPUs in R

### 2022-01-04 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
p1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3087 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
p1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3089 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

<<setup, cache=FALSE,include=FALSE>>=
library('knitr')
options(continue="+  ",  prompt="R> ", digits=3, width=57)

@@

\Abstract{
We introduce the \proglang{R} package \pkg{clrng} which leverages the \pkg{gpuR} package to generate random numbers on a Graphics Processing Unit (GPU) with the \pkg{clRNG} (OpenCL) library. Parallel processing with GPU's can speed up computationally intensive tasks, which when combined with \proglang{R}, it can largely improve \proglang{R}’s downsides in terms of slow speed, memory usage and computation mode.  %There is currently no \proglang{R} package that does random number generation on GPU, while random number generation is critical in simulation-based statistical inference and modeling. 
\pkg{clrng} enables reproducible research by setting random streams on GPU and can thus accelerate several types of simulation and modelling. This package is portable and flexible, developers can use its random number generation kernel for various other purposes and applications. 
}

@@

The \pkg{clrng} package described here is currently the only \proglang{R} package that is able to generate random numbers on GPUs. Accomplishing this is complicated because each process must produce an independent (non-overlapping) sequence of random numbers, and in order to ensure reproducibility it should be possible to save and restore the current state of each stream of random numbers at any point.  Here we introduce the \proglang{R} package \pkg{clrng} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by using the \proglang{clRNG} \citep{l2015clrng} library in \proglang{OpenCL}, and can thus accelerate several types of statistical simulation and modelling.

@@

%The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item copies the current states only of the streams to its private memory. 
\fct{setBaseCreator} sets the initial state of the first stream that is created, it plays a role in \pkg{clrng} like the function \fct{SetPackageSeed} in \pkg{RngStreams} and \fct{clrngSetBaseCreatorState} in \pkg{clRNG}. \pkg{clrng} is able to create streams both on the host and on the GPU device. The \fct{createStreamsCpu} function does the former. The \proglang{R} output below creates 4 streams on the host. 
%is an interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on host as \proglang{R} matrices, see the \proglang{R} output below, where we created 4 streams on the host using \fct{createStreamsCpu}, after transposing \code{myStreamsCpu}, each column of the matrix is a stream, creating streams on host makes it easy for us to view and check the streams behavior. Substreams are not created in \pkg{clrng} because currently we use one stream per work-item. 
<<createStreams_CPU, message=FALSE>>=
# creating streams on CPU
library("clrng")
setBaseCreator(rep(12345,6))
myStreamsCpu <- createStreamsCpu(4)
t(myStreamsCpu)
@
\fct{setBaseCreator} has one argument:
\begin{itemize}
\itemsep0em 
  \item \code{initial}: a vector of six 31-bit integers, recycled if length is shorter. Default is set to \code{c(12345, 12345, 12345, 12345, 12345, 12345)}.
\end{itemize}
\fct{createStreamsCpu} has one argument:
\begin{itemize}
\itemsep0em 
  \item \code{n}: number of streams to create.
\end{itemize}
We move streams to the GPU by converting them to a `\code{vclMatrix}'.
<<createStreams_Gpu, eval=FALSE>>=
myStreamsGpu = vclMatrix(myStreamsCpu)
@
Equivalently, \fct{createStreamsGpu} creates streams directly on the GPU and returns a `\code{vclMatrix}', which makes it slightly more efficient when the number of streams is large.
<<createStreams_Gpu2>>=
myStreamsGpu2 = createStreamsGpu(4)
@
%Users should run either \fct{createStreamsCpu} or \fct{createStreamsGpu} at the beginning of a task and retain these streams for all of the random number generation the task requires.  Creating new streams with different initial states introduces the very small possibility of overlap between the original and new streams. If an \proglang{R} session is restarted during a task, perhaps because a preliminary analysis is done following an initial set of simulations and the full set of simulations continued at a later date, the state of the streams should be transferred to the CPU and saved to disk.  The `\code{initial}' argument to the stream creator should be used with caution and we recommend using the default.  It bears repeating that two sets of streams with different initial states may produce overlapping numbers, the probability is small but increases when the number of streams is large.
There are a few important notes on the usage of these above functions. (1), We allow users to call \fct{setBaseCreator}  at any time. \fct{setBaseCreator} should be called once before \fct{createStreamsCpu} or \fct{createStreamsGpu} if they would like to select their own initial seeds, or never, in which case streams take the package's default initial seed. (2), Users may switching between \fct{createStreamsCpu} and \fct{createStreamsGpu} in the same program, however, this is not recommended as it causes unnecessary data transfer between host and device. (3), If an \proglang{R} session is restarted during a task, as long as users saved the \proglang{R} environment, the initial states (seeds) of the newly created streams will carry on from the last streams' seed from the previous \proglang{R} session.

@@

Now we use the 4 streams created in section \ref{createstreams} to generate a vector of double-precision i.i.d. $U(0,1)$ random numbers with \fct{clrng::runif}. To view the generated random numbers, we need to convert them to \proglang{R} vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
<<simStuff2, eval=TRUE,tidy=TRUE, echo=TRUE>>=
as.vector(clrng::runif(n=6, streams=myStreamsGpu2, Nglobal=c(2,2)))
@
The arguments of the \fct{clrng::runif} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: A vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: Streams for random number generation, streams cannot be missing.
  \item \code{Nglobal}: Global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
  \item \code{verbose}: Print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}
%\fct{runif} returns the same sequence of random numbers because it uses the same streams \code{myStreamsGpu} and \code{myStreamsGpu2} here. 
Reusing \code{myStreamsGpu} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu} after it being used.
<<eval=FALSE,tidy=TRUE, include=FALSE>>=
t(matrix(as.matrix(myStreamsGpu), nrow(myStreamsCpu), ncol(myStreamsCpu), dimnames = dimnames(myStreamsCpu)))
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. There are two ways to reproduce results, depending on how the program calls streams creator. The simpler way to reproduce is just to restart the program with the same initial seed for the creator.  But it is important that the program always creates the streams in exactly the same order, for example, like the toy \proglang{R} program for demonstration below, we can reproduce the random matrix \code{sim_mat} just by keeping a record of the initial seed \code{c(11,22,33,44,55,66)}.
<<showreproduce1, echo=TRUE, eval=FALSE>>=
library('clrng')
setBaseCreator(c(11,22,33,44,55,66))
sim_mat <- matrix(0, nrow=10, ncol= 6)
for (i in 1:10){
  if( i %% 2 == 0){
    streams1 <- createStreamsGpu(4)
    sim_mat[i,] <- as.vector(clrng::rnorm(n=6, streams=streams1, Nglobal=c(2,2)))
  }else if(i==3){
    streams2 <- createStreamsGpu(4)
    sim_mat[i,] <- as.vector(clrng::runif(n=6, streams=streams2, Nglobal=c(2,2)))
  }else{
    streams3 <- createStreamsGpu(8)
    sim_mat[i,] <- as.vector(clrng::rexp(n=6, streams=streams3, Nglobal=c(4,2)))
}
}
sim_mat
@
For more complicated applications, we recommend users to save the matrix of streams (current states and initial states) on the CPU in a file as a \code{.rds} object. And later recall the saved streams for regenerating results, the streams will start from their current states after they are loaded. This is a safer way than the previous one for reproducing results in simulations. Below shows the code that saves streams to a data file called \code{myStreams.rds} on CPU and then load it back and transfer it to a `vclMatrix' \code{streams_saved} on GPU, and using it to (re)generate some Normal random numbers.
<<save streams,echo=TRUE,eval=FALSE>>=
saveRDS(as.matrix(myStreamsGpu2), "myStreams.rds")
# Load the streams object as streams_saved
streams_saved <- vclMatrix(readRDS("myStreams.rds"))
clrng::rnorm(n=6, streams=streams_saved, Nglobal=c(2,2))
@

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is above 160 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
<<random normal time compare, eval=TRUE, cache=FALSE>>=
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, Nglobal=c(512,128), type="double"))
@
<<random normal time compare2, eval=TRUE, cache=TRUE>>=
system.time(matrix(stats::rnorm(10000^2),10000,10000))
@

@@

 <<monthdata,eval=TRUE,echo=FALSE, cache=TRUE, fig.pos='h', message=FALSE,purl=FALSE>>=
library(knitr)
library(tidyverse)
library(kableExtra)
knitr::kable(month,format="latex", align = c("rrrrrrrrrrrr"), #label="tab:month"
  caption = "Monthly birth anomaly data\\label{tab:month}", booktabs=TRUE,linesep = "") %>%
  kable_styling(full_width = F, position = "center")#latex_options = "HOLD_position")
@
<<monthDataForScript, include=FALSE, purl=TRUE>>-
knitr::kable(month)
@

>>>>

<<setup, cache=FALSE,include=FALSE>>=
library('knitr')
options(continue="+  ",  prompt="R> ", digits=3, width=58)

@@

\Abstract{
We introduce the \proglang{R} package \pkg{clrng} which leverages the \pkg{gpuR} package and is able to generate random numbers in parallel on a Graphics Processing Unit (GPU) with the \pkg{clRNG} (OpenCL) library.  Parallel processing with GPU's can speed up computationally intensive tasks, which when combined with \proglang{R}, it can largely improve \proglang{R}’s downsides in terms of slow speed, memory usage and computation mode.  
%There is currently no \proglang{R} package that does random number generation on GPU, while random number generation is critical in simulation-based statistical inference and modeling. 
\pkg{clrng} enables reproducible research by setting random initial seeds for streams on GPU and CPU, and can thus accelerate several types of statistical simulation and modelling. 
The random number generator in \pkg{clrng} guarantees independent parallel samples even when \proglang{R} is used interactively in an ad-hoc manner, with sessions being interrupted and restored. This package is portable and flexible, developers can use its random number generation kernel for various other purposes and applications.  
}

@@

The \pkg{clrng} package described here is currently the only \proglang{R} package that provides facilities for generating random numbers in parallel on GPU. Accomplishing this is complicated because each process must produce an independent (non-overlapping) sequence of random numbers, and in order to ensure reproducibility it should be possible to save and restore the current state of each stream of random numbers at any point.  By leveraging the \pkg{gpuR} package, \pkg{clrng} provides an extensible framework for \proglang{R} package developers to make use of these facilities in \proglang{C} or \proglang{OpenCL} code, and the random number generators guarantee independent parallel samples even when \proglang{R} is used interactively in an ad-hoc manner, with sessions being interrupted and restored.
%and can generate random numbers on GPU by using the \proglang{clRNG} \citep{l2015clrng} library in \proglang{OpenCL}, and can thus accelerate several types of statistical simulation and modelling.

@@

%The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item copies the current states only of the streams to its private memory. 
\fct{setBaseCreator} sets the initial state of the first stream that is created, it plays a role in \pkg{clrng} like the function \fct{SetPackageSeed} in \pkg{RngStreams} and \fct{clrngSetBaseCreatorState} in \pkg{clRNG}. \pkg{clrng} is able to create streams both on the host and on the GPU device. The \fct{createStreamsCpu} function does the former. The \proglang{R} output below creates 4 streams on the host. 
%is an interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on host as \proglang{R} matrices, see the \proglang{R} output below, where we created 4 streams on the host using \fct{createStreamsCpu}, after transposing \code{myStreamsCpu}, each column of the matrix is a stream, creating streams on host makes it easy for us to view and check the streams behavior. Substreams are not created in \pkg{clrng} because currently we use one stream per work-item. 
<<createStreams_CPU, message=FALSE>>=
# creating streams on CPU
library("clrng")
setBaseCreator(rep(12345,6))
myStreamsCpu <- createStreamsCpu(4)
t(myStreamsCpu)
@
\fct{setBaseCreator} has one argument:
\begin{itemize}
\itemsep0em 
  \item \code{initial}: a vector of six 31-bit integers, recycled if length is shorter. Default is set to \code{c(12345, 12345, 12345, 12345, 12345, 12345)}.
\end{itemize}
\fct{createStreamsCpu} has one argument:
\begin{itemize}
\itemsep0em 
  \item \code{n}: number of streams to create.
\end{itemize}
We move streams to the GPU by converting them to a `\code{vclMatrix}'.
<<createStreams_Gpu, eval=TRUE>>=
myStreamsGpu = vclMatrix(myStreamsCpu)
@
Equivalently, \fct{createStreamsGpu} creates streams directly on the GPU and returns a `\code{vclMatrix}', which makes it slightly more efficient when the number of streams is large.
<<createStreams_Gpu2>>=
myStreamsGpu2 = createStreamsGpu(4)
@
%Users should run either \fct{createStreamsCpu} or \fct{createStreamsGpu} at the beginning of a task and retain these streams for all of the random number generation the task requires.  Creating new streams with different initial states introduces the very small possibility of overlap between the original and new streams. If an \proglang{R} session is restarted during a task, perhaps because a preliminary analysis is done following an initial set of simulations and the full set of simulations continued at a later date, the state of the streams should be transferred to the CPU and saved to disk.  The `\code{initial}' argument to the stream creator should be used with caution and we recommend using the default.  It bears repeating that two sets of streams with different initial states may produce overlapping numbers, the probability is small but increases when the number of streams is large.
There are a few important notes on the usage of these above functions. (1), We allow users to call \fct{setBaseCreator}  at any time. \fct{setBaseCreator} should be called once before \fct{createStreamsCpu} or \fct{createStreamsGpu} if they would like to select their own initial seeds, or never, in which case streams take the package's default initial seed. (2), Users may switching between \fct{createStreamsCpu} and \fct{createStreamsGpu} in the same program, however, this is not recommended as it causes unnecessary data transfer between host and device. (3), If an \proglang{R} session is restarted during a task, as long as users saved the \proglang{R} environment, the initial states (seeds) of the newly created streams will carry on from the last streams' seed from the previous \proglang{R} session.

@@

Now we use the 4 streams created in section \ref{createstreams} to generate a vector of double-precision i.i.d. $U(0,1)$ random numbers with \fct{clrng::runif}. To view the generated random numbers, we need to convert them to \proglang{R} vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
<<simStuff2, eval=TRUE,tidy=TRUE, echo=TRUE>>=
as.vector(clrng::runif(n=6, streams=myStreamsGpu2, Nglobal=c(2,2)))
@
The arguments of the \fct{clrng::runif} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: A vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: Streams for random number generation, streams cannot be missing.
  \item \code{Nglobal}: Global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
  \item \code{verbose}: Print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}
%\fct{runif} returns the same sequence of random numbers because it uses the same streams \code{myStreamsGpu} and \code{myStreamsGpu2} here. 
Reusing \code{myStreamsGpu} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu} after it being used.
<<eval=FALSE,tidy=TRUE, include=FALSE>>=
t(matrix(as.matrix(myStreamsGpu2), nrow(myStreamsCpu), ncol(myStreamsCpu), dimnames = dimnames(myStreamsCpu)))
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. There are two ways to reproduce results, depending on how the program calls streams creator. The simpler way to reproduce is just to restart the program with the same initial seed for the creator.  But it is important that the program always creates the streams in exactly the same order, for example, like the toy \proglang{R} program for demonstration below, we can reproduce the random matrix \code{sim_mat} just by keeping a record of the initial seed \code{c(11,22,33,44,55,66)}.
<<showreproduce1, echo=TRUE, eval=FALSE>>=
library('clrng')
setBaseCreator(c(11,22,33,44,55,66))
sim_mat <- matrix(0, nrow=10, ncol= 6)
for (i in 1:10){
  if( i %% 2 == 0){
    streams1 <- createStreamsGpu(4)
    sim_mat[i,] <- as.vector(clrng::rnorm(n=6, streams=streams1, Nglobal=c(2,2)))
  }else if(i==3){
    streams2 <- createStreamsGpu(4)
    sim_mat[i,] <- as.vector(clrng::runif(n=6, streams=streams2, Nglobal=c(2,2)))
  }else{
    streams3 <- createStreamsGpu(8)
    sim_mat[i,] <- as.vector(clrng::rexp(n=6, streams=streams3, Nglobal=c(4,2)))
}
}
sim_mat
@
For more complicated applications, we recommend users to save the matrix of streams (current states and initial states) on the CPU in a file as a \code{.rds} object. And later recall the saved streams for regenerating results, the streams will start from their current states after they are loaded. This is a safer way than the previous one for reproducing results in simulations. Below shows the code that saves streams to a data file called \code{myStreams.rds} on CPU and then load it back and transfer it to a `vclMatrix' \code{streams_saved} on GPU, and using it to (re)generate some Normal random numbers.
<<save streams,echo=TRUE,eval=FALSE>>=
saveRDS(as.matrix(myStreamsGpu2), "myStreams.rds")
# Load the streams object as streams_saved
streams_saved <- vclMatrix(readRDS("myStreams.rds"))
clrng::rnorm(n=6, streams=streams_saved, Nglobal=c(2,2))
@

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is above 160 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
<<random normal time compare, eval=TRUE, cache=FALSE>>=
streams <- createStreamsGpu(512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, Nglobal=c(512,128), type="double"))
@
<<random normal time compare2, eval=TRUE, cache=TRUE>>=
system.time(matrix(stats::rnorm(10000^2),10000,10000))
@

@@

 <<monthdata,eval=TRUE,echo=FALSE, cache=TRUE, fig.pos='h', message=FALSE,purl=FALSE>>=
library(knitr)
library(tidyverse)
library(kableExtra)
knitr::kable(month,format="latex", align = c("rrrrrrrrrrrr"), #label="tab:month"
  caption = "Monthly birth anomaly data\\label{tab:month}", booktabs=TRUE,linesep = "") %>%
  kable_styling(full_width = F, position = "center")#latex_options = "HOLD_position")
@
<<monthDataForScript, include=FALSE, purl=TRUE>>=
knitr::kable(month)
@
